Very simplified likelihood

### 2024-09-03 inst/very_simplified_likelihood.Rmd
fixes #240

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $R$, the rank of the focal competitor: we
have $1\leqslant R\leqslant n+1$.  Alternatively we might consider
$r=R-1$, the number of $b$ clones finishing ahead of $a$, so
$0\leqslant r\leqslant n$.  The initial field strength is
$a+nb=1+b(n-1)$.

$$r=R-1$$

Suppose our bro comes first (zero $b$'s ahead of him) [$r=0, R=1$]:

\[
\mathcal{L}(a)=\frac{a}{a+nb}=\frac{1-b}{1+(n-1)b}
\]

Comes second [$r=1,R=2$], one $b$ ahead of him:

\[
\mathcal{L}(a)=\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
\]

Third [$r=2, R=3$]; two ahead of him:

\[
\mathcal{L}(a)=
\frac{b}{a+ n   b}\cdot
\frac{b}{a+(n-1)b}\cdot
\frac{a}{a+(n-2)b}
\]

Fourth [$r=3, R=4$], three ahead:

\[
\mathcal{L}(a)=
\frac{b}{a+ n   b}\cdot
\frac{b}{a+(n-1)b}\cdot
\frac{b}{a+(n-2)b}\cdot
\frac{a}{a+(n-3)b}
\]

Placing after $r<n$ of the $n$ clones of strength $b$ finish:

\[
\mathcal{L}(a)=
\frac{b^ra}{
(a+nb)(a+(n-1)b)(a+(n-2)b)\cdots(a+(n-r)b)}
=
\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}
\]

where $B=1/b$ [also using the fact that $a+b=1$].

This is

\[
\mathcal{L}_{r,n}(a)=\frac{(B-1)(B+n-r-2)!}{(B+n-1)!},\qquad B=1/(1-a)
\]

\[
\mathcal{L}_{R,n}(a)=\frac{(B-1)(B+n-R-1)!}{(B+n-1)!},\qquad B=1/(1-a)
\]

\[
=\frac{(B-1)\Gamma(B+n-R)}{\Gamma(B+n)},\qquad B=1/(1-a)
\]

Then, using Stirling [viz $n!\sim n^ne^{-n}\sqrt{2\pi n}$] we get

\[
\log(B-1) + ((B+n-r-2)\log(B+n-r-2)-(B+n-r-2)+\frac{\log(B+n-r-2)}{2})
          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
\]

\[
\log(B-1) + ((B+n-R-1)\log(B+n-R-1)-(B+n-R-1)+\frac{\log(B+n-R-1)}{2})
          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
\]

If you give Sage this:

```
var('B n r')
pretty_print_default(True)
X=log(B-1) + (
+ ((B+n-r-2)*log(B+n-r-2)-(B+n-r-2)+log(B+n-r-2)/2)
- ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
)
X.taylor(n,Infinity,2)
```

It returns this:

\[
-(r+1)\log n
-\frac{(2B-3)r-r^2+2B-2}{2n}
-\frac{3(2B-3)r^2-6B^2-6(B^2-3B+2)r+12B-5}{12n^2}
+\log(B-1)+\mathcal{O}(n^{-3})
\]

Now using $R$ not $r$:

```
sage: var('B n R')
(B, n, R)
pretty_print_default(False)
X = log(B-1) + (
  + ((B+n-R-1)*log(B+n-R-1)-(B+n-R-1)+log(B+n-R-1)/2)
  - ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
  )
X
(B - R + n - 1)*log(B - R + n - 1) - (B + n - 1)*log(B + n - 1) + R + 1/2*log(B - R + n - 1) - 1/2*log(B + n - 1) + log(B - 1)
X.taylor(n,Infinity,2).full_simplify()
-1/12*(12*R*n^2*log(n) + 3*(2*B - 1)*R^2 - 2*R^3 - 12*n^2*log(B - 1) - 6*(B^2 - B)*R + 6*((2*B - 1)*R - R^2)*n)/n^2
```

Note that if we use instead $n~\sim n^ne^{-n}\sqrt{2\pi
n}\left(1+\frac{1}{12n}\right)$, then only terms of
$\mathcal{O}(n^{-2})$ and above are changed; the first-order terms and
indeed the zeroth order terms are unaffected.  Anyway, after
simplifying and taking an asymptotic expansion to order $n^{-1}$, we
find, to first order,

\[
\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+ K + \mathcal{O}\left(n^{-2}\right)
\]

(where $K$ is an arbitrary constant).  Observing that
$\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate would
be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
results below.

@@

```
f[B_, n_, R_] := Log[(B - 1)*Gamma[B + n - R]/Gamma[B + n]] 
Assuming[B > 1 && p > 0 && p < 1, 
 Series[f[B, n, n*p], {n, Infinity, 1}]]
```

We get

$$
\log\left(
(B-1)(1-p)^{B-1/2} +
\frac{
	(B-1)(1-6B+6B^2)(1-p)^{B-1/2}p
}{
	12(1-p)n
}
\right)
$$

or

$$
\log\left(
(B-1)(1-p)^{B-1/2} \left[1+
\frac{(1-6B+6B^2)p}{12(1-p)n}\right]\right)
$$

$$
\log
(B-1) + B\log(1-p) +
n^{-1}
\frac{(1-6B+6B^2)p}{12(1-p)} +{\mathcal O}\left(n^{-2}\right)
$$

$$
\log
(B-1) + B\log(1-p) +
\frac{1}{12}n^{-1}\frac{p}{1-p}(1-6B+6B^2) + {\mathcal O}\left(n^{-2}\right)
$$

# Placing last

Placing last is finishing after $n=r$ clones finish, that is, $R=n+1$
[me plus two clones, I come third].

\[
n=1,R=2\longrightarrow\mathcal{L}=b\]

\[
n=2,R=3\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
=\frac{1}{B(B+1)}
\]

\[
n=3,R=4\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
\]

\[
n=4,R=n+1\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
\frac{b^4}{(1+3b)(1+2b)(1+b)}=
\frac{1}{B(B+1)(B+2)(B+3)}
\]

For $n=4$ we have equivalently

\[
\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+2)}
\]

In general we would have

\[
\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
\frac{(B-1)!}{(B+n-1)!}=
\frac{\Gamma(B)}{\Gamma(B+n)}
\]

(denominator has $n+1$ terms).  Observe that this agrees with the
$r<n$ case above (substituting $r=n$).  We may introduce some R idiom:

```{r,label=definefsingle}
f_single <- function(a,R,n,log=FALSE){  # not vectorised
    stopifnot(R>0)
	B <- 1/(1-a)
	if(log){
		out <- log(B-1) + lgamma(B+n-R) - lgamma(B+n)
	} else {  # not-log
		out <- (B-1)/prod(B+(n-R):(n-1))
	}
   return(out)
}

f_vec_a <- function(a,R,n, log=FALSE){  # vectorised in 'a' but not in 'R'
	sapply(a,function(a){f_single(a,R=R,n=n,log=log)})
}

f_vec_arn <- function(a,R,n,log=FALSE){
	## vectorized in 'a'; treats 'R' and 'n' as
	## vectors of independent observations

	M <- cbind(R,n)

	if(log){
		out <- 0
		for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,R=M[i,1],n=M[i,2],log=TRUE)}
	} else {
		out <- 1
		for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,R=M[i,1],n=M[i,2],log=FALSE)}
	}
	return(out)
}

fapprox <- function(a,R,n,log=FALSE){
	if(log){
		return(log(B-1)-B*R/n)
	} else {
		return((B-1)/exp(R*B/n))
	}
}
```

```{r label=defdiff}
fdash <- function(a,R,n,log=TRUE){
	B <- 1/(1-a)
	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-R))*B^2
	if(!log){out <- out * f_vec_arn(a,R,n,log=FALSE)}
	return(out)
}
```

@@

```{r label=verifydiff}
a <- 0.34
d <- 1e-3
R <- 4
n <- 10
c(
  numerical  = (f_single(a+d/2,R,n,log=TRUE)-f_single(a-d/2,R,n,log=TRUE))/d,
  analytical = fdash(a,R,n)
)

c(
  numerical  = (f_single(a+d/2,R,n,log=FALSE)-f_single(a-d/2,R,n,log=FALSE))/d,
  analytical = fdash(a,R,n,log=FALSE)
)
```

@@

```{r,label=defmle}
MLE <- function(R,n,give=FALSE){
	d <- 1e-6
	out <- optimize(f_vec_arn,c(d,1-d),R=R,n=n,log=TRUE,maximum=TRUE)
	if(!give){out <- out$maximum}
	return(out)
}

MLEa <- function(R,n,give=FALSE,small=1e-4){
	out <- uniroot(f=function(a){fdash(a,R,n)},interval=c(small,1-small))
	if(!give){out <- out$root}
	return(out)
}

@@

```{r usestirling}
rbind(
showdiff(MLE(10, 100), 100 / 111),
showdiff(MLE(10, 500), 500 / 511),
showdiff(MLE(10,1000),1000 /1011),
showdiff(MLE(10,5000),5000 /5011)
)
```

```{r usestirlinga}
rbind(
showdiff(MLEa(10, 100), 100 / 111),
showdiff(MLEa(10, 500), 500 / 511),
showdiff(MLEa(10,1000),1000 /1011),
showdiff(MLEa(10,5000),5000 /5011)
)

@@

```{r,label=verifytwo}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9
R <- 4
c(
way1 = b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
way2 = (B-1)/prod(B+(5:8)),
way3 = f_vec_arn(a,R,n,log=FALSE)
)
```

```{r,label=makejj}
out <- list()
for(n in 1:10){
  out[[n]]  <- rep(NA,n+1)
  for(R in seq_len(n+1)){
    if(R==1){
      jj <- 1
    } else if(R == n+1){
      jj <- 0
    } else {
      jj <- MLE(R,n)
    }
    out[[n]][R] <- jj
  }
}
out
```

```{r,label=plotjj}
plotterp <- function(...){
plot(NA,xlim=c(1,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
for(n in 1:10){
  for(R in seq_len(n+1)){   points(n,out[[n]][R],pch=16) }
  for(R in seq_len(n+1)){   text(n+0.2,out[[n]][R],R,cex=0.5,col='gray') }
}
}
plotterp()
pdf(file="dotprobs.pdf")
plotterp()
dev.off()
```

@@

The likelihood calculations above can be confusing becuase likelihood
is not the same as probability.

Consider, for example, $n=2$ so we have 3 competitors.

probability of focal competitor coming first [i.e. $r=0$]:

@@

Now what is the probability of coming second, that is, $r=1$?

@@

\[
\mbox{prob of coming second}   =
\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad\mbox{WRONG}
\]

but this is incorrect: there are two clones of (strength) $b$, and the
likelihood arguments above do not distinguish between their finishing
order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
means that the finishing order was $b_1\succ a\succ b_2$ or $b_2\succ
a\succ b_1$.  Noting that these events are disjoint, the probability
would be

@@

Similarly for the focal competitor coming last we need to sum the
probabilities of finishing orders $b_1\succ b_2\succ a$ and $b_2\succ
b_1\succ a$:

@@

Now try $r=3$ [i.e. 4 competitors].  Taking things one step at a time:

@@

Now consider the general case.  Again, $n$ clones of strength $b$ and
one of strength $a=1-b$, $n+1$ competitors.

@@

Now consider the case where there are $r$ clones coming ahead of the
focal competitor:

\[
\mbox{prob of having $r$ clones ahead}
=\frac{n!}{(n-r)!}\cdot X
\]

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 8
jj <- f_vec_arn(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
grid()
rain <- rainbow(n+1)
for(R in seq_len(n+1)){
  y <- f_vec_a(a,R,n,log=FALSE)
  if(R==1){y[length(y)] <- 1}
  y <- y/max(y,na.rm=TRUE)
  if(R>1){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rain[R])
}
abline(v=MLE(R=5,n=8))
text(0.07,0.95,"R=9",col=rain[9])
text(0.20,0.95,"R=8",col=rain[8])
text(0.26,0.91,"R=7",col=rain[7])
text(0.29,0.83,"R=6",col=rain[6])
text(0.33,0.76,"R=5",col=rain[5])
text(0.35,0.64,"R=4",col=rain[4])
text(0.39,0.53,"R=3",col=rain[3])
text(0.45,0.41,"R=2",col=rain[2])
text(0.54,0.21,"R=1",col=rain[1])
}

@@

```{r shown4n8mle}
MLE(R=5,n=8)
```

@@

d_ocon_max  <- MLE(d_ocon ,22)
d_gasly_max <- MLE(d_gasly,22)
f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
f_vec_arn(bothmax,d_gasly,22,log=TRUE)
f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)

Lambda <- (
   (f_vec_arn(d_gasly_max  ,d_ocon,22,log=TRUE)+f_vec_arn(d_gasly_max  ,d_gasly,22,log=TRUE)) -
   (f_vec_arn(bothmax      ,d_ocon,22,log=TRUE)+f_vec_arn(bothmax      ,d_gasly,22,log=TRUE))
)
Lambda
pchisq(-2*Lambda,df=1,lower.tail=FALSE)
```

@@

```{r checofirst}
checo_like <- function(a){
  out <- a*0
  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year),getnum(year)-1,log=TRUE) }
  return(out)
}

@@

```{r definef1log}
f1_logistic <- function(vec){
  alpha <- vec[1]
  beta  <- vec[2]
  out <- 0
  for(year in 2011:2023){
     LO <- alpha + beta*(year-2011)
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,getfoc(year),getnum(year)-1,log=TRUE)
  }
  return(out)
}
```

@@

  d1max <- MLE(d1,n)
  d2max <- MLE(d2,n)
  Lambda <- (
  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
)
  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
}

@@

```{r analyze_athletics_olymp,cache=TRUE}
jjf <- function(x){f_vec_arn(x,O$rank,O$n-1,log=TRUE)}
system.time(optolym <- optimize(jjf, c(0.45,0.65), maximum=TRUE))
```

@@

```{r make_athletic_like}
a <- seq(from=0.35,to=0.65,len=45)
L <- f_vec_arn(a,O$rank,O$n-1,log=TRUE)
Lmax <- jjf(optolym$maximum)
Lmax
L <- L-Lmax
```

@@

```{r calculatelikes1,cache=F}
a <- seq(from=0.005,to=0.85,by=0.01)
La <- f_vec_arn(a,R=results$place[1],results$runners[1]-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
plot(a,La,type='b',ylab='fishy')
abline(h=c(0,-2))
```

```{r splitresults}
supp2 <- function(vec,place,runners){
  a_stirling <- vec[1]
  a_auckland <- vec[2]
  out <- 0
  for(i in 1:3){
    out <- out + f_vec_a(a_stirling,R=place[i],runners[i]-1,log=TRUE)
  }
  for(i in 4:10){
    out <- out + f_vec_a(a_auckland,R=place[i],runners[i]-1,log=TRUE)
  }
  return(out)
}
```

@@

```{r logisticsupportparkrun,cache=FALSE}
results <- read.table("parkrun_results.txt",header=TRUE)
results
supp_logistic <- function(vec){
  alpha <- vec[1]
  beta <- vec[2]
  out <- 0
  for(i in seq_len(nrow(results))){
     LO <- alpha + beta*i
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,results$place[i]-1,results$runners[i]-1,log=TRUE)
  }
  return(out+13400)
}

@@

```{r dopurefurst}
Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure,ylab='pure strength')
optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
```

```{r definesupp2ed}
supp2_ed <- function(vec,place1,place2,runners1,runners2){
  out <- 0
  M1 <- cbind(place1,runners1)
  M2 <- cbind(place2,runners2)
  for(i in seq_len(nrow(M1))){
    out <- out + f_vec_arn(vec[1],R=M1[i,1],n=M1[i,2]-1,log=TRUE)
  }
  for(i in seq_len(nrow(M2))){
    out <- out + f_vec_arn(vec[2],R=M2[i,1],n=M2[i,2]-1,log=TRUE)
  }
  return(out)
}
```

>>>>

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n-1$ competitors each of strength $b$; we require $a+b=1$.  NB we
have $n$ competitors in total, 1 of strength $a$ and $n-1$ of strength
$b$.  An observation is indexed by $r$, the rank of the focal
competitor: we have $1\leqslant r\leqslant n$.  The initial field
strength is $a+(n-1)b=1+b(n-2)b$.

Suppose our bro comes first (zero $b$'s ahead of him) [$r=1$]:

\[
\mathcal{L}(a)=\frac{a}{a+(n-1)b}=\frac{1-b}{1+(n-2)b}
\]

Comes second [$r=2$], one $b$ ahead of him:

\[
\mathcal{L}(a)=\frac{b}{a+(n-1)b}\cdot\frac{a}{a+(n-2)b}=
\frac{b(1-b)}{(1+(n-2)b)\cdot(1+(n-3)b)}
\]

Third [$r=3$]; two ahead of him:

\begin{eqnarray}
\mathcal{L}(a) &=&
\frac{b}{a+ n   b}\cdot
\frac{b}{a+(n-1)b}\cdot
\frac{a}{a+(n-2)b}\\
&=&
\frac{b^2(1-b)\rule{0mm}{10mm}
}{
(1+(n-2)b)\cdot(1+(n-3)b)\cdot(1+(n-4)b)}
\end{eqnarray}

Fourth [$r=4$], three clones finish before the focal competitor:

\[
\mathcal{L}(a)=
\frac{b^3(1-b)
}{
(1+(n-2)b)\cdot(1+(n-3)b)\cdot(1+(n-4)b)\cdot(1+(n-5)b)}
\]

Placing with general rank $r$, $1\leqslant r\leqslant n$.  Rank $r$
means $r-1$ clones finish ahead of the focal competitor:

\begin{eqnarray}
\mathcal{L}(a) &=&
\frac{b^{r-1}(1-b)
}{
\underbrace{(1+(n-2)b)\cdot(1+(n-3)b)\cdots(1+(n-r-1)b)}_{\mbox{$r$ terms}}
}
\\&=&
\frac{B-1\rule{1mm}{10mm}}{(n+B-2)\cdot(n+B-3)\cdots(n+B-r-1)}
\end{eqnarray}

where $B=1/b$.  Recall that $r$ is the rank and $r-1$ clones finish
ahead of the FC.  This is

\[
\mathcal{L}_{r,n}(a)=\frac{(B-1)(n+B-r-2)!}{(n+B-2)!},\qquad B=1/(1-a)
\]

\[
=\frac{(B-1)\Gamma(n+B-r-1)}{\Gamma(n+B-1)},\qquad B=1/(1-a)
\]

@@

```
f[B_, n_, r_] := 
 Log[(B - 1)] + Log[Gamma[n + B - r - 1]/Gamma[n + B - 1]]
Assuming[B > 1 && p > 0 && p < 1, 
 Series[f[B, n, n*p], {n, Infinity, 1}]]
```

gives

\[
\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1) + (B-3/2)\log(1-r/n) + 
\mathcal{O}\left(n^{-1}\right)
\]

(where $K$ is an arbitrary constant).  Here, $\mathcal{S}$ is the
support (log-likelihood).  Observing that
$\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate would
be unique.

# R idiom

```{r, label=definefsingle}
f_single <- function(a, r, n) {  # not vectorised
    stopifnot(r>0)
B <- 1/(1-a)
  log(B-1) + lgamma(n+B-r-1) - lgamma(n+B-1)
}

f_vec_a <- function(a,r,n){  # vectorised in 'a' but not in 'R'
	sapply(a,function(a){f_single(a,r=r,n=n)})
}

f_vec_arn <- function(a,r,n){
	## vectorized in 'a'; treats 'r' and 'n' as
	## vectors of independent observations

	M <- cbind(r,n)

    out <- 0
    for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,r=M[i,1],n=M[i,2])}
	return(out)
}

fapprox <- function(a,r,n){
    stop("not written")
}
```

```{r label=defdiff}
fdash <- function(a,r,n){  # dS/da = dS/dB*dB/da
	B <- 1/(1-a)
    (1/(B-1) + psigamma(n+B-r-1) - psigamma(n+B-1))*B^2
}
```

@@

```{r label=verifydiff}
a <- 0.34
d <- 1e-3
r <- 4
n <- 10
c(
  numerical  = (f_single(a+d/2, r, n) - f_single(a-d/2, r, n))/d,
  analytical = fdash(a,r,n)
)

@@

```{r,label=defmle}
MLE <- function(r,n,give=FALSE){
	d <- 1e-6
	out <- optimize(f_vec_arn, c(d, 1-d), r=r, n=n, maximum=TRUE)
	if(!give){out <- out$maximum}
	return(out)
}

MLEa <- function(r,n,give=FALSE,small=1e-4){
	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(small,1-small))
	if(!give){out <- out$root}
	return(out)
}

@@

@@

```{r,label=verifytwo}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9  # 9 runners
r <- 4  # FC came 4th
c(
way1 = b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
way2 = (B-1)/prod(B+(5:8)),
way3 = f_vec_arn(a,r,n)
)
```

```{r,label=makejj}
out <- list()
for(n in 1:10){
  out[[n]]  <- rep(NA,n+1)
  for(r in seq_len(n+1)){
    if(r==1){
      jj <- 1
    } else if(r == n+1){
      jj <- 0
    } else {
      jj <- MLE(r,n)
    }
    out[[n]][r] <- jj
  }
}
out
```

```{r,label=plotjj}
plotterp <- function(...){
plot(NA,xlim=c(2,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
for(n in 2:10){
  for(r in seq_len(n)){ points(n,out[[n]][r],pch=16,col='black') }
  for(r in seq_len(n)){ text(n+0.2,out[[n]][r],r,cex=0.5,col='gray') }
}
}
plotterp()
pdf(file="dotprobs.pdf")
plotterp()
dev.off()
```

@@

The likelihood calculations above can be confusing becuase likelihood
is not the same as probability.  Consider, for example, $n=3$ so we
have 3 competitors: the focal competitor of strength $a$ and two
clones each of strength $b=1-a$.

probability of focal competitor coming first [i.e. $r=1$]:

@@

Now what is the probability of coming second, that is, $r=2$?

@@

\[
\mbox{prob of coming second}   =
\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad{\mbox{WARNING}}
\]

but this is *incorrect*: there are two clones of strength $b$, and the
likelihood arguments above do not distinguish between their finishing
order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
that $b_1$ and $b_2$ both have strength $b$.  Then "the focal
competitor came second" means that the finishing order was $b_1\succ
a\succ b_2$ or $b_2\succ a\succ b_1$.  Noting that these events are
disjoint, the probability would be

@@

[see the factor of two difference].  Similarly for the focal
competitor coming last we need to sum the probabilities of finishing
orders $b_1\succ b_2\succ a$ and $b_2\succ b_1\succ a$:

@@

Now try $n=4$ [i.e. 4 competitors: the FC and three clones].  Taking
things one step at a time:

@@

Now consider the general case.  This time we consider $n$ clones of
strength $b$ and one of strength $a=1-b$, $n+1$ competitors.

@@

Now consider the case where there are $p$ clones coming ahead of the
focal competitor:

\[
\mbox{prob of having $r$ clones ahead}
=\frac{n!}{(n-p)!}\cdot X
\]

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 9 # the FC and 8 clones
jj <- exp(f_vec_arn(a,3,n))
plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
grid()
rain <- rainbow(n)
for(r in seq_len(n)){
  y <- exp(f_vec_a(a,r,n))
  if(r==1){y[length(y)] <- 1}
  y <- y/max(y,na.rm=TRUE)
  if(r>1){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rain[r])
}
abline(v=MLE(r=5,n=8))
text(0.09,0.95,"r=9",col=rain[9])
text(0.18,0.93,"r=8",col=rain[8])
text(0.25,0.90,"r=7",col=rain[7])
text(0.29,0.83,"r=6",col=rain[6])
text(0.33,0.76,"r=5",col=rain[5])
text(0.35,0.65,"r=4",col=rain[4])
text(0.39,0.54,"r=3",col=rain[3])
text(0.45,0.41,"r=2",col=rain[2])
text(0.59,0.21,"r=1",col=rain[1])
}

@@

```{r shown4n8mle}
MLE(r=5,n=8)
```

@@

d_ocon_max  <- MLE(d_ocon ,22)
d_gasly_max <- MLE(d_gasly,22)
f_vec_arn(bothmax,d_ocon ,22)
f_vec_arn(bothmax,d_gasly,22)
f_vec_arn(d_ocon_max ,d_ocon ,22)
f_vec_arn(d_gasly_max,d_gasly,22)

Lambda <- (
   (f_vec_arn(d_gasly_max  ,d_ocon,22)+f_vec_arn(d_gasly_max  ,d_gasly,22)) -
   (f_vec_arn(bothmax      ,d_ocon,22)+f_vec_arn(bothmax      ,d_gasly,22))
)
Lambda
pchisq(-2*Lambda,df=1,lower.tail=FALSE)
```

@@

```{r checofirst}
checo_like <- function(a){
  out <- a*0
  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year),getnum(year)) }
  return(out)
}

@@

```{r definef1log}
f1_logistic <- function(vec){
  alpha <- vec[1]
  beta  <- vec[2]
  out <- 0
  for(year in 2011:2023){
     LO <- alpha + beta*(year-2011)
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,getfoc(year),getnum(year))
  }
  return(out)
}
```

@@

  d1max <- MLE(d1,n)
  d2max <- MLE(d2,n)
  Lambda <- (
  (f_vec_arn(d2max  ,d1,n) + f_vec_arn(d2max  ,d2,n)) -
  (f_vec_arn(bothmax,d1,n) + f_vec_arn(bothmax,d2,n))
)
  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
}

@@

```{r analyze_athletics_olymp,cache=TRUE}
jjf <- function(x){f_vec_arn(x,O$rank,O$n)}
system.time(optolym <- optimize(jjf, c(0.45,0.65), maximum=TRUE))
```

@@

```{r make_athletic_like}
a <- seq(from=0.35,to=0.65,len=45)
L <- f_vec_arn(a,O$rank,O$n)
Lmax <- jjf(optolym$maximum)
Lmax
L <- L-Lmax
```

@@

```{r calculatelikes1, cache=F}
a <- seq(from=0.005,to=0.85,by=0.01)
La <- f_vec_arn(a,r=results$place[1],results$runners[1])
La <- La - max(La,na.rm=TRUE)
plot(a,La,type='b',ylab='fishy')
abline(h=c(0,-2))
```

```{r splitresults}
supp2 <- function(vec,place,runners){
  a_stirling <- vec[1]
  a_auckland <- vec[2]
  out <- 0
  for(i in 1:3){
    out <- out + f_vec_a(a_stirling,r=place[i],runners[i])
  }
  for(i in 4:10){
    out <- out + f_vec_a(a_auckland,r=place[i],runners[i])
  }
  return(out)
}
```

@@

```{r logisticsupportparkrun,cache=FALSE}
results <- read.table("parkrun_results.txt",header=TRUE)
results
supp_logistic <- function(vec){
  alpha <- vec[1]
  beta <- vec[2]
  out <- 0
  for(i in seq_len(nrow(results))){
     LO <- alpha + beta*i
     strength <- exp(LO)/(1+exp(LO))
     out <- out + f_vec_arn(strength,results$place[i],results$runners[i])
  }
  return(out+13400)
}

@@

```{r dopurefurst}
Lpure <- f_vec_arn(a,rank[wp],class_size[wp])
Lpure <- Lpure - max(Lpure,na.rm=TRUE)
plot(a,Lpure,ylab='pure strength')
optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp])},c(0.3,0.6),maximum=TRUE)
optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa])},c(0.3,0.6),maximum=TRUE)
```

```{r definesupp2ed}
supp2_ed <- function(vec,place1,place2,runners1,runners2){
  out <- 0
  M1 <- cbind(place1,runners1)
  M2 <- cbind(place2,runners2)
  for(i in seq_len(nrow(M1))){
    out <- out + f_vec_arn(vec[1],r=M1[i,1],n=M1[i,2])
  }
  for(i in seq_len(nrow(M2))){
    out <- out + f_vec_arn(vec[2],r=M2[i,1],n=M2[i,2])
  }
  return(out)
}
```
